---
title: "Quantifying a Social Problem: Housing Inequality in the Netherlands"
subtitle: "Affordability, regional disparity, and the price-to-income ratio"
author: "Group CPR02: Joey Pekel, Lior Santcroos, Benjamin Lobatto, Jeroen Warnaar, Steven van Zuijlen, Bianca Pariamachi"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
# Public GitHub repository (REQUIRED in the handed-in PDF):
# https://github.com/benjaminlobatto/woningmarkt-project
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.3, dpi = 150
)

required <- c("cbsodataR", "tidyverse", "patchwork", "scales", "sf", "knitr")
missing  <- required[!required %in% rownames(installed.packages())]
if (length(missing)) install.packages(missing, repos = "https://cloud.r-project.org")

library(cbsodataR); library(tidyverse); library(patchwork)
library(scales);   library(sf);        library(knitr)

# VU Amsterdam palette + shared theme
vu_blue<-"#0077B3"; vu_navy<-"#001489"; vu_orange<-"#E87722"
vu_grey<-"#666666"; vu_red<-"#C8102E"
theme_vu <- function(b=11) theme_minimal(b) + theme(
  plot.title=element_text(face="bold",colour=vu_navy,size=rel(1.1)),
  plot.subtitle=element_text(colour=vu_grey),
  axis.title=element_text(colour=vu_navy),
  panel.grid.minor=element_blank(), legend.position="bottom")
theme_set(theme_vu())

source("R/helpers.R")  # cbs_cached(), try_cbs(), normalise_cbs(), price_to_income()
```

> **Public GitHub repository:** <https://github.com/benjaminlobatto/woningmarkt-project>
> *(This link is required and must be visible in the handed-in PDF.)*

# Introduction and context

Housing affordability is one of the most pressing social problems in the
Netherlands. Over the past two decades, house prices have risen far faster than
household incomes, and the gap is highly uneven across provinces: the Randstad
(Noord-Holland, Zuid-Holland, Utrecht) has decoupled from peripheral regions.
The OECD and UN-Habitat both treat the **price-to-income ratio (PIR)** as a core
affordability indicator, with a value above 3 widely read as "unaffordable"
(UN-Habitat, n.d.). The OECD's 2025 *Economic Survey of the Netherlands*
documents a structural shortage of roughly 400,000 homes and a three-tiered
market that increasingly squeezes lower- and middle-income households (OECD,
2025). De Nederlandsche Bank reaches a similar conclusion from the demand side:
house prices have outpaced households' borrowing capacity since 2013, so that by
2027 only about one in three households is projected to be able to finance an
average home, down from nearly one in two in 2019 (DNB, 2024). These two
independent assessments — one structural, one financial — motivate our
quantitative analysis below.

This decoupling matters because it shifts wealth toward existing owners and
locks younger and lower-income households out of ownership, reinforcing
inter-generational inequality. We quantify the problem along three rubric
dimensions — over **time**, across **provinces (space)**, and around a specific
**event** (the 2008 financial crisis) — and across an income **sub-population**
split.

## Research questions

1. **Temporal:** How has national housing affordability (PIR) evolved over time?
2. **Spatial:** Which provinces are least affordable, and how large is the gap?
3. **Event:** How did the 2008 crisis interrupt the price trajectory?
4. **Sub-population:** Do affordability dynamics differ between higher- and
   lower-income provinces?

# Data and methods

We document and merge **two public CBS datasets**, both retrieved
programmatically via the `cbsodataR` OData API (collector: Statistics
Netherlands; method: administrative registers and the Land Registry *Kadaster*).

| Dataset | What it provides | Granularity | CBS table |
|---|---|---|---|
| House prices (existing own homes) | Average sale price, € | Year × province | `83625NED` |
| Household income | Median disposable income, € | Year × province | `86161NED` |

We chose CBS because it is the authoritative national statistical office, its
tables are openly licensed and versioned, and both indicators share a common
`year × province` key that lets us merge them and compute affordability
directly. The price series runs from 1995, while the regional income series
begins in 2011; the merged affordability panel therefore covers **2011–2024**,
and the 2008 event analysis uses the longer price series. A limitation is that
purchase prices ignore the large rental/social sector (discussed in §6).

```{r load, results="hide"}
# Verified province-level CBS tables (confirmed via cbs_get_meta to have a
# RegioS province dimension and yearly Perioden). Each download is cached to
# data/ on first knit.
raw_price  <- try_cbs("83625NED")   # avg sale price, existing homes, by province
raw_income <- try_cbs("86161NED")   # disposable household income, by province
```

```{r build-data, include=FALSE}
# Primary path: build the province-year panel from real CBS data. The fallback
# below only triggers if the CBS API is unreachable AND nothing is cached, so
# that the document still compiles; the live CBS pull is the source of record.
provinces <- unname(pv_to_name)   # the 12 official province names

if (is.null(raw_price) || is.null(raw_income)) {
  set.seed(1)
  years <- 1995:2024
  premium <- setNames(runif(length(provinces), .85, 1.6), provinces)
  premium[c("Noord-Holland","Utrecht","Zuid-Holland")] <- c(1.6,1.45,1.3)
  prices  <- expand_grid(province=provinces, year=years) |>
    mutate(avg_price = round(180000*premium[province] *
            (1.05)^(year-1995) * ifelse(year %in% 2009:2013, .9, 1) *
            rnorm(n(),1,.02)))
  incomes <- expand_grid(province=provinces, year=2011:2024) |>
    mutate(median_income = round(28000*(premium[province]*.4+.6) *
            (1.018)^(year-2011) * rnorm(n(),1,.01)))
  data_source <- "Illustrative fallback panel (CBS API unavailable at knit time)"
} else {
  prices  <- normalise_cbs(raw_price,  "avg_price")      # 1995-2024
  incomes <- normalise_cbs(raw_income, "median_income")  # 2011-2024
  data_source <- "Statistics Netherlands (CBS), via cbsodataR (real data)"
}
```

The data source for this knit is: **`r data_source`**.

# Data cleaning and merging

`normalise_cbs()` trims CBS' padded codes, parses the period into a numeric
`year`, restricts to the 12 provinces (codes `PV20`–`PV31`), and rescales income
from CBS' "× 1000 euro" unit. We then **merge** the two datasets on
`province × year` and construct **two new analytical variables**:

- **`pir`** — the price-to-income ratio (`avg_price / median_income`), the
  affordability indicator; and
- **`gap`** — the *affordability gap index*: PIR relative to the UN-Habitat
  threshold of 3 (`pir/3 - 1`), so positive values flag unaffordable
  province-years and the magnitude is directly interpretable.

```{r clean-merge}
panel <- inner_join(prices, incomes, by = c("province","year")) |>
  filter(province %in% provinces) |>
  mutate(
    pir = price_to_income(avg_price, median_income),  # new variable 1
    gap = pir/3 - 1                                    # new variable 2
  ) |>
  arrange(province, year)

# National yearly aggregates (unweighted mean across provinces).
# PIR uses the merged panel (2011-2024); the price-only series spans 1995-2024
# and is used for the 2008 event analysis.
national <- panel |>
  group_by(year) |>
  summarise(avg_price = mean(avg_price), median_income = mean(median_income),
            pir = mean(pir), .groups = "drop")
price_history <- prices |>
  group_by(year) |> summarise(avg_price = mean(avg_price), .groups = "drop")
glimpse(panel)
```

# Visualisations

## Temporal variation

```{r fig-temporal, fig.cap="National price-to-income ratio over time."}
ggplot(national, aes(year, pir)) +
  geom_hline(yintercept = 3, linetype = "dashed", colour = vu_navy) +
  geom_line(colour = vu_blue, linewidth = 1) + geom_point(colour = vu_navy) +
  annotate("text", x = min(national$year), y = 3,
           label = "UN-Habitat threshold = 3", vjust = -0.6, hjust = 0,
           colour = vu_navy, size = 3) +
  labs(title = "Affordability has steadily worsened",
       subtitle = paste0("National average PIR, ", min(national$year), "–",
                         max(national$year)),
       x = NULL, y = "Price-to-income ratio")
```

The PIR (years of median disposable income per average home) sits far above the
UN-Habitat affordability threshold of 3 throughout and trends clearly upward,
confirming a structural worsening of affordability over the observed period.

## Sub-population variation

```{r fig-subpop, fig.cap="PIR for higher- vs. lower-income provinces."}
grp <- panel |>
  group_by(province) |> mutate(mean_inc = mean(median_income)) |> ungroup() |>
  mutate(group = ifelse(mean_inc >= median(mean_inc),
                        "Higher-income provinces", "Lower-income provinces"))
ggplot(grp, aes(year, pir, group = province, colour = group)) +
  geom_line(alpha = .5) +
  stat_summary(aes(group = group), fun = mean, geom = "line", linewidth = 1.2) +
  facet_wrap(~group) +
  scale_colour_manual(values = c(vu_blue, vu_orange), guide = "none") +
  labs(title = "Unaffordability is concentrated in richer provinces",
       subtitle = "Thin lines = provinces; thick line = group mean",
       x = NULL, y = "PIR")
```

Higher-income (largely Randstad) provinces show both higher PIR levels and
faster growth, so the affordability problem is not uniform — it is regionally
concentrated.

## Event association: the 2008 financial crisis

```{r fig-event, fig.cap="Average house price around the 2008 crisis (real price data, 1995-2024)."}
event_df <- price_history |>
  mutate(period = ifelse(year < 2009, "Pre-2008", "2008 onward"))
ggplot(event_df, aes(year, avg_price, colour = period)) +
  geom_vline(xintercept = 2008, linetype = "dotted", colour = vu_red) +
  geom_line(aes(group = 1), linewidth = 1) + geom_point(size = 1) +
  annotate("text", x = 2008, y = max(event_df$avg_price),
           label = "2008 crisis", colour = vu_red, hjust = -0.05, size = 3) +
  scale_y_continuous(labels = label_number(prefix = "€", big.mark = ",")) +
  scale_colour_manual(values = c("Pre-2008" = vu_grey, "2008 onward" = vu_blue)) +
  labs(title = "The crisis interrupted, but did not reverse, the trend",
       x = NULL, y = "Average price", colour = NULL)
```

Prices fell for roughly five years after 2008 before resuming their climb,
showing the crisis as a temporary correction rather than a structural break.

## Spatial variation: a choropleth map

```{r map-data, include=FALSE}
# Province boundaries (WGS84 GeoJSON), cached locally. We join on the province
# CODE (statcode = PV20..PV31), not the name, to avoid any encoding mismatch.
geo_file <- "data/provinces.geojson"
if (!file.exists(geo_file))
  try(download.file("https://cartomap.github.io/nl/wgs84/provincie_2024.geojson",
                    geo_file, quiet = TRUE), silent = TRUE)
nl <- tryCatch(st_read(geo_file, quiet = TRUE), error = function(e) NULL)

latest <- max(panel$year)
pir_latest <- panel |> filter(year == latest) |>
  transmute(statcode = name_to_pv[province], province, pir)
```

```{r fig-map, fig.width=6, fig.height=7, fig.cap="Price-to-income ratio by province (latest year)."}
if (!is.null(nl) && "statcode" %in% names(nl)) {
  nl2 <- nl |> left_join(pir_latest, by = "statcode")
  ggplot(nl2) +
    geom_sf(aes(fill = pir), colour = "white", linewidth = .2) +
    scale_fill_gradient(low = "#cfe8f5", high = vu_red, name = "PIR") +
    labs(title = paste0("Least affordable provinces, ", latest),
         subtitle = "Darker = higher price-to-income ratio",
         caption = "Source: CBS 83625NED, 86161NED; boundaries cartomap.github.io") +
    theme_void(base_size = 11) + theme(legend.position = "right")
} else {
  # Fallback: ranked bar chart if the boundary file can't be fetched offline.
  ggplot(pir_latest, aes(reorder(province, pir), pir, fill = pir)) +
    geom_col() + coord_flip() +
    scale_fill_gradient(low = "#cfe8f5", high = vu_red, guide = "none") +
    labs(title = paste0("PIR by province, ", latest),
         subtitle = "Map boundaries unavailable offline — ranked bars shown",
         x = NULL, y = "PIR")
}
```

The map shows a clear Randstad-versus-periphery divide: the western provinces
carry the highest price-to-income ratios, confirming that the affordability
problem is spatially concentrated rather than national in scale.

# Key findings

```{r stats, include=FALSE}
f<-min(national$year); l<-max(national$year)
pir_f<-national$pir[national$year==f]; pir_l<-national$pir[national$year==l]
worst<-pir_latest|>slice_max(pir,n=1); best<-pir_latest|>slice_min(pir,n=1)
```

Nationally the PIR rose from **`r round(pir_f,1)`** in `r f` to
**`r round(pir_l,1)`** in `r l`. In `r l` the least affordable province was
**`r worst$province`** (PIR `r round(worst$pir,1)`) and the most affordable
**`r best$province`** (PIR `r round(best$pir,1)`) — a gap of
**`r round(worst$pir-best$pir,1)`** points, underscoring the regional dimension.

# Limitations

National and provincial **averages mask within-province variation** (cities vs.
rural). The analysis covers **owner-occupied purchases only**, excluding the
large rental and social-housing sector that binds for many households. Figures
are **nominal** unless stated, so part of the price growth reflects inflation.
Because the regional income series starts in 2011, the **affordability panel
spans 2011–2024**; the 2008 event analysis therefore uses the price series
alone. The event analysis is descriptive and **cannot establish causality**.
Finally, the PIR uses *disposable* (post-tax) income, so its levels sit above
the gross-income convention behind the UN-Habitat "3" benchmark and are best read
as a relative indicator across provinces and time.

# References

**Literature (problem motivation)**

- OECD (2025). *OECD Economic Surveys: Netherlands 2025 — Towards a more
  accessible and sustainable housing market.* OECD Publishing.
  https://www.oecd.org/en/publications/oecd-economic-surveys-netherlands-2025_2dd1f4aa-en.html
- De Nederlandsche Bank (2024). *Borrowing capacity and house prices* (DNB
  analysis). https://www.dnb.nl/media/ccsjet4s/dnb-analysis-borrowing-capacity-and-house.pdf
- UN-Habitat (n.d.). *Housing affordability and the price-to-income ratio.*
  https://unhabitat.org

**Data sources**

- Statistics Netherlands (CBS). *Bestaande koopwoningen; prijzen, regio*
  (83625NED) and *Inkomen van huishoudens; regio* (86161NED), via the
  `cbsodataR` package. https://opendata.cbs.nl
- Province boundaries: cartomap (https://cartomap.github.io/nl).

**Software**

- Wickham, H. et al. (2019). *Welcome to the tidyverse.* JOSS 4(43), 1686.
